Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-04-08 Manuscript/manuscript.Rmd
manuscript edits

Commonly used measures of model performance for binary classifiers include threshold-free metrics like the area under the receiver operating curve (AUROC), and other threshold-specific metrics like accuracy, sensitivity, and specificity. The ROC curve is created by plotting the true positive rate (TPR) against the false positive rate (FPR) at various threshold settings. The subsequent AUROC, measures how well the model is capable of distinguishing between the two classes [@park2004receiver]. Accuracy is defined as the ratio of correctly identified cases (either from the positive or negative class) over the total number of cases. It measures the overall proportion of correctly classified instances. Sensitivity, also referred to as recall or TPR, is the ratio of true positives over the total number of positive classes. Specificity, on the other hand, also referred to as TNR, is the ratio of true negatives over the total number of negative classes. Each measures the proportion of correctly identified positive and negative classes respectively [@yerushalmy1947statistical]. Problems arise when using such measures in the event of imbalanced classes. For AUROC, when the prevalence of an event is low--that is, a defined genomic region truly containing a TAD boundary--mean probabilities are biased towards the majority class. This results in the area of interest under an ROC curve to be compressed toward a small corner in the ROC space [@prati2011survey]. While not directly affecting the value of the AUROC, this can have severe consequences on the interpretability of the model's performance. A better threshold-free measure to use when dealing with imbalanced data is the area under the precision-recall curve (AUPRC).  The interpretation of the AUPRC of a model is similar to that of the AUROC. However, where as the baseline model performance according to the AUROC is given by 0.50, the baseline model performance according to the AUPRC is given by the ratio of true positives over the total number of cases. This enables a more realistic benchmark to evaluate a model on as this ratio will be lower than 0.5 for a test data set with imbalanced classes. Likewise, threshold-specific metrics like accuracy and specificity are also problematic when evaluating model performance on imbalance data due to being highly influenced by the correct classification of the majority class. Better measures to consider are composite metrics involving both precision and recall together because they soften the influence of a model's ability to correctly classify majority classes and yield more dependable and interpretable results. Therefore, measures like the F1-Score and Matthew's Correlation Coefficient (MCC) offer better interpretation, especially in the instance of imbalanced classes [@jeni2013facing]. The F1-Score can be interpreted as a weighted average of the precision and recall values, while the MCC is, in essence, the correlation coefficient between the observed and predicted binary classifications [@brodersen2010balanced]. Each metric uses all 4 quadrants of the confusion matrix, thus providing more balanced measures in the case of imbalanced data [@chicco2017ten].

@@

The efficacy of each re-sampling technique was compared using composite metrics inclusing F1-score and MCC. These metrics are defined as follows:

@@

Here, TP refers to the number of bins correctly identified as containing a TAD boundary (true positives), FP refers to the number of bins incorrectly identified as containing a TAD boundary (false positives), TN refers to the number of bins correctly identified as not containing a TAD boundary (true negatives), and FN refers to the number of bins incorrectly identified as not containing a TAD boundary. Each of these quantities are obtained from the confusion matrix created by validating the model on the testing set. Additional threshold-free metrics where also reported including the AUROC and AUPRC.

Lastly, for the task of identifying the genomic annotations that were most predictive of the formation of TAD boundaries, we evaluated the variable importances associated with each feature of the random forest model associated with the best combination of reduced predictor type and re-sampling technique, across each cell line. To calculate variable importances for each model, the AUPRC on the out-of-bag portion of the data within each tree was recorded. Then the same was done after permuting each predictor variable. The difference between the two AUPRC values was then averaged over all trees, and normalized by the standard error of the differences. All measures of importances were then rescaled to have a maximum value of 100 for better interpretation.

@@

Lastly, we investigated which genomic annotations were most predictive of TAD boundary formation. Looking at heatmaps of variable importances provided by the random forest classifiers implemented on SMOTE re-sampled data, with distance-type predictors, we found that known transcription factors associated with the hypothesized loop extrusion model [@fudenberg2016formation; @alipour2012self] were give high predictive importance in the model (Figure 10). These tended to include the SMC3ab, ZNF143, CTCF, and RAD21 transcription factor binding sites and were found to be ranked consistly across resolution. Other annotations known to be associated with gene regulation and affect structural properties of chromatin, such as the YY1 and MAZ transcription factors, as well DNAse hypersensitive sites, regions of heterochromatin, and the H3k27 trimethylation and H3K4me1 modifications were less consistently ranked across resolutions for either cell line.

>>>>

Commonly used measures of model performance for binary classifiers include threshold-free metrics like the area under the receiver operating curve (AUROC), and other threshold-specific metrics like accuracy, sensitivity, and specificity. The ROC curve is created by plotting the true positive rate (TPR) against the false positive rate (FPR) at various threshold settings. The subsequent AUROC, measures how well the model is capable of distinguishing between the two classes [@park2004receiver]. Accuracy is defined as the ratio of correctly identified cases (either from the positive or negative class) over the total number of cases. It measures the overall proportion of correctly classified instances. Sensitivity, also referred to as recall or TPR, is the ratio of true positives over the total number of positive classes. Specificity, on the other hand, also referred to as TNR, is the ratio of true negatives over the total number of negative classes. Each measures the proportion of correctly identified positive and negative classes respectively [@yerushalmy1947statistical]. Problems arise when using such measures in the event of imbalanced classes. For AUROC, when the prevalence of an event is low--that is, a defined genomic region truly containing a TAD boundary--mean probabilities are biased towards the majority class. This results in the area of interest under an ROC curve to be compressed toward a small corner in the ROC space [@prati2011survey]. While not directly affecting the value of the AUROC, this can have severe consequences on the interpretability of the model's performance. <!--A better threshold-free measure to use when dealing with imbalanced data is the area under the precision-recall curve (AUPRC).  The interpretation of the AUPRC of a model is similar to that of the AUROC. However, where as the baseline model performance according to the AUROC is given by 0.50, the baseline model performance according to the AUPRC is given by the ratio of true positives over the total number of cases. This enables a more realistic benchmark to evaluate a model on as this ratio will be lower than 0.5 for a test data set with imbalanced classes.--> Likewise, threshold-specific metrics like accuracy and specificity are also problematic when evaluating model performance on imbalance data due to being highly influenced by the correct classification of the majority class. Better measures to consider are composite metrics involving both precision and recall together because they soften the influence of a model's ability to correctly classify majority classes and yield more dependable and interpretable results. Therefore, measures like the F1-Score and Matthew's Correlation Coefficient (MCC) offer better interpretation, especially in the instance of imbalanced classes [@jeni2013facing]. The F1-Score can be interpreted as a weighted average of the precision and recall values, while the MCC is, in essence, the correlation coefficient between the observed and predicted binary classifications [@brodersen2010balanced]. Each metric uses all 4 quadrants of the confusion matrix, thus providing more balanced measures in the case of imbalanced data [@chicco2017ten].

@@

The efficacy of each re-sampling technique was compared using composite metrics including F1-score and MCC. These metrics are defined as follows:

@@

Here, TP refers to the number of bins correctly identified as containing a TAD boundary (true positives), FP refers to the number of bins incorrectly identified as containing a TAD boundary (false positives), TN refers to the number of bins correctly identified as not containing a TAD boundary (true negatives), and FN refers to the number of bins incorrectly identified as not containing a TAD boundary. Each of these quantities are obtained from the confusion matrix created by validating the model on the testing set. <!--Additional threshold-free metrics where also reported including the AUROC and AUPRC.--> Lastly, for the task of identifying the genomic annotations that were most predictive of the formation of TAD boundaries, we ranked the variable importances associated with each feature of the ensemble system-specific random forest model. <!--To calculate variable importances for each model, the F1-Score on the out-of-bag portion of the data within each tree was recorded. Then the same was done after permuting each predictor variable. The difference between the two AUPRC values was then averaged over all trees, and normalized by the standard error of the differences.--> Annotations with low rankings correspond to functional epigenomic features that are either positively or negatively enriched near TAD boundaries, and thus, better allow the model to discriminate between genomic bins that overlap with flanked TAD boundaries.

@@

Lastly, we investigated which genomic annotations were most predictive of TAD boundary formation. Looking at heatmaps of variable importances provided by the random forest classifiers implemented on SMOTE re-sampled data, with distance-type predictors, we found that known transcription factors associated with the hypothesized loop extrusion model [@fudenberg2016formation; @alipour2012self] were give high predictive importance in the model for both cell lines (Figure 6 A-B). These tended to include the SMC3ab, ZNF143, CTCF, and RAD21 transcription factor binding sites and were found to be ranked consistly across resolution. Other annotations known to be associated with gene regulation and affect structural properties of chromatin, such as the YY1 and MAZ transcription factors, as well DNAse hypersensitive sites, regions of heterochromatin, and the H3k27 trimethylation and H3K4me1 modifications were less consistently ranked across resolutions for either cell line. Moreover, the ranking profiles of genomic annotations were found to be more similar between the 10 kb and 25 kb resolutions, as well as between the 50 kb and 100 kb resoluions.
